%! suppress = EscapeAmpersand


\section{QUIC overview}
\label{sec:quic-overview}
QUIC is a connection-based, multistream protocol that uses UDP under the hood.
It is integrated with TLS and provides own congestion control and flow control mechanisms.
Data sent over QUIC is delivered in a reliable, ordered way.
Together with QUIC there comes a new version of HTTP called HTTP/3.

This section outlines main features of QUIC and compares QUIC to the other transport protocols.

\subsection{Packet types in QUIC}
\label{subsec:quic-packet-types}
The most outer entity in QUIC is a packet.
Each QUIC packet consists of a header that can take long or short form and one or more QUIC frames.
QUIC packets are directly encapsulated into UDP datagrams.
This is illustrated in figure~\ref{fig:quic-packet-format}.
A single UDP datagram can also coalesce many QUIC packets, however, this feature is not important in terms of regular data exchange.

\begin{figure}
    \centering
    \begin{bytefield}{32}
        \begin{leftwordgroup}{\tiny UDP header}
            \bitbox{16}{\tiny source port} & \bitbox{16}{\tiny destination port} \\
            \bitbox{16}{\tiny length} & \bitbox{16}{\tiny checksum}
        \end{leftwordgroup} \\
        \begin{leftwordgroup}{\tiny UDP data}
            \begin{rightwordgroup}{\tiny QUIC packet}
                \wordbox{1}{\tiny quic header} \\
                \wordbox[tlr]{1}{\tiny  quic frames} \\
                \wordbox[blr]{1}{\tiny $\cdots$}
            \end{rightwordgroup}
        \end{leftwordgroup}
    \end{bytefield}
    \caption{QUIC packet format}
    \label{fig:quic-packet-format}
\end{figure}

QUIC defines two types of header: short and long.
Figure~\ref{fig:short-header-format} illustrates Short Header format.
The first byte encodes miscellaneous flags.
Next, there is a destination connection ID that takes at most 20 bytes and a packet number that takes from 1 to 4 bytes.
Total size of Short Header is from 2 to 25 bytes.
Fields marked with red color are subject to an encryption process.
On the other hand Long Header additionally includes version, destination connection ID length, source connection ID length and source connection ID\@.
Flags differ a little comparing to the Short Header but together with a packet number length they also take 1 byte.

\begin{figure}
    \centering
    \begin{bytefield}[bitwidth=4em]{8}
        \bitheader{0-7} \\
        \bitbox{1}{\tiny header form} & \bitbox{1}{\tiny fixed bit} & \bitbox{1}{\tiny spin bit} & \bitbox{2}[bgcolor=lightred]{\tiny reserved bits} & \bitbox{1}[bgcolor=lightred]{\tiny key phase} & \bitbox{2}[bgcolor=lightred]{\tiny packet number length} \\
        \wordbox{1}{\tiny destination connection ID (0 - 160)} \\
        \wordbox{1}[bgcolor=lightred]{\tiny packet number (8 - 32)}
    \end{bytefield}
    \caption{Short Header format}
    \label{fig:short-header-format}
\end{figure}

QUIC introduces also different types of packets and frames.
Depending on specific packet type, it can contain different frame types.
For example \textit{Handshake} packet can contain \textit{CRYPTO}, \textit{PING}, \textit{PADDING}, \textit{ACK} or \textit{CONNECTION\_CLOSE} frames.
We can divide packet types basing on header form they use.
Following packet types use Long Header: \textit{Version Negotiation}, \textit{Initial}, \textit{0-RTT}, \textit{Handshake}, \textit{Retry}.
They are sent at the beginning of communication and therefore they are not covered more here.

Packet that uses Short Header is called \textit{1-RTT}.
1-RTT packets are sent after version and 1-RTT keys negotiation.
In an interactive communication they are main packets and they will, for example, convey media data.
1-RTT packet format is illustrated in figure~\ref{fig:1rtt-packet-format}.

\begin{figure}
    \centering
    \begin{bytefield}[bitwidth=3.5em]{8}
        \bitheader{0-7} \\
        \begin{leftwordgroup}{\tiny short \\ \tiny header}
            \bitbox{1}{\tiny header form} & \bitbox{1}{\tiny fixed bit} & \bitbox{1}{\tiny spin bit} & \bitbox{2}[bgcolor=lightred]{\tiny reserved bits} & \bitbox{1}[bgcolor=lightred]{\tiny key phase} & \bitbox{2}[bgcolor=lightred]{\tiny packet number length} \\
            \wordbox{1}{\tiny destination connection ID (0 - 160)} \\
            \wordbox{1}[bgcolor=lightred]{\tiny packet number (8 - 32)}
        \end{leftwordgroup} \\
        \wordbox[tlr]{1}[bgcolor=lightred]{\tiny payload} \\
        \wordbox[blr]{1}[bgcolor=lightred]{\tiny $\cdots$}
    \end{bytefield}
    \caption{1-RTT packet format}
    \label{fig:1rtt-packet-format}
\end{figure}

\subsection{Security and low-latency connection establishment}
\label{subsec:security-and-low-latency-connection-establishment}
QUIC is integrated with TLS 1.3 providing confidentiality, integrity and authenticity of its packets.
Integration with TLS 1.3 is described in details in RFC 9001~\cite{rfc9001}.
It is based on negotiating both transport and cryptographic parameters in a single connection handshake which results in a low-latency connection establishment.
Figure~\ref{fig:low-latency-conn-est} compares time needed for sending HTTP request using TCP/TLS and QUIC.

\begin{figure}
    \centering
    \begin{subfigure}{.5\textwidth}
        \begin{sequencediagram}
            \newinst{client}{Client}
            \newinst[3]{server}{Server}
            \mess{client}{TCP SYN}{server}
            \mess{server}{TCP SYN + ACK}{client}
            \mess{client}{TCP ACK}{server}
            \postlevel
            \mess{client}{TLS ClientHello}{server}
            \mess{server}{TLS ServerHello}{client}
            \mess{client}{TLS Finished}{server}
            \postlevel
            \mess{client}{HTTP REQ}{server}
            \mess{server}{HTTP RES}{client}
        \end{sequencediagram}
        \caption{HTTP request in TCP}
        \label{subfig:http-req-tcp}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \begin{sequencediagram}
            \newinst{client}{Client}
            \newinst[3]{server}{Server}
            \mess{client}{QUIC}{server}
            \mess{server}{QUIC}{client}
            \mess{client}{QUIC}{server}
            \postlevel
            \mess{client}{HTTP REQ}{server}
            \mess{server}{HTTP RES}{client}
        \end{sequencediagram}
        \caption{HTTP request in QUIC}
        \label{subfig:http-req-quic}
    \end{subfigure}
    \caption{HTTP request comparison}
    \label{fig:low-latency-conn-est}
\end{figure}

Round Trip Time (RTT) is a time needed for transporting a message from one side to the other and back again.
Standard HTTP request using TCP/TLS stack requires 3-RTT -- one for TCP handshake, one for TLS handshake and one for HTTP request.
Even though TCP handshake sends 3 messages it is still counted as 1-RTT because we do not have to wait for a response to the last message but we can immediately send next data, which in this case is data for TLS handshake.

QUIC takes another approach.
It combines transport and cryptographic handshakes reducing time needed for establishing a connection.
In such a case sending an HTTP request occurs after 1-RTT\@.
The whole communication lasts 2-RTT\@.

However, QUIC can reduce time needed for sending an HTTP request and receiving a response even more.
In case when there was communication with the server previously and client cached information from it, it is possible to send 0-RTT HTTP request.
This scenario is presented in figure~\ref{fig:http-req-quic-0rtt}
\begin{figure}
    \centering
    \begin{sequencediagram}
        \newinst{client}{Client}
        \newinst[3]{server}{Server}
        \mess{client}{QUIC}{server}
        \mess{client}{HTTP REQ}{server}
        \postlevel
        \mess{server}{QUIC}{client}
        \mess{server}{HTTP RES}{client}
        \postlevel
        \mess{client}{QUIC}{server}
    \end{sequencediagram}
    \caption{HTTP request in QUIC with 0-RTT packets}
    \label{fig:http-req-quic-0rtt}
\end{figure}

The only limitation of integrating TLS with QUIC is that there is no possibility to send unencrypted data.

\subsection{Stream multiplexing}
\label{subsec:stream-multiplexing}
QUIC is a multiplexed protocol which means we can open many streams in one QUIC connection.
Stream is an ordered sequence of bytes.
Each stream can be unidirectional or bidirectional and can be opened by a client or by a server.
Different types of streams can have different flow control limits (see section~\ref{subsec:flow-control-and-congestion-control}).
Data in specific streams is carried independently which means that losing packet in one stream does not affect flow in other streams.
Figure~\ref{fig:stream-multiplexing} shows this scenario.
One of the packets in stream 1 (marked as gray) was lost and needs to be retransmitted blocking subsequent packets (marked as black) on the receiver side.
They cannot be conveyed to the application layer until retransmission and reordering of the lost packet.
This is the source of so called head of line blocking problem which appears when we are trying to multiplex many HTTP requests in a single TCP connection.
Losing one packet stops the entire flow.
However, thanks to the stream multiplexing QUIC resolves this problem and streams 2 and 3 are not affected by the packet loss in stream 1.

\begin{figure}
    \centering
    \begin{tikzpicture}

        \node[cylinder,
        draw = violet,
        text = purple,
        style={transform shape},
        cylinder uses custom fill,
        cylinder body fill = magenta!10,
        cylinder end fill = magenta!40,
        minimum size = 3.5cm] (c) at (0,0) {QUIC connection};

        \draw [->, postaction={decorate,decoration={raise=2ex, text along path,text align=center,text={stream 1}}}] (2, 1.5) -- (7.5, 1.5);
        \filldraw [fill=green!20, draw=black] (2.5, 1.25) rectangle (3.25, 1.75);
        \filldraw [fill=gray!20, draw=black] (3.75, 1.25) rectangle (4.5, 1.75);
        \filldraw [fill=black!60, draw=black] (5, 1.25) rectangle (5.75, 1.75);
        \filldraw [fill=black!60, draw=black] (6.25, 1.25) rectangle (7, 1.75);

        \draw [->, postaction={decorate,decoration={raise=2ex, text along path,text align=center,text={stream 2}}}] (2, 0) -- (7.5, 0);
        \filldraw [fill=cyan!20, draw=black] (2.5, -0.25) rectangle (3.25, 0.25);
        \filldraw [fill=cyan!20, draw=black] (3.75, -0.25) rectangle (4.5, 0.25);
        \filldraw [fill=cyan!20, draw=black] (5, -0.25) rectangle (5.75, 0.25);
        \filldraw [fill=cyan!20, draw=black] (6.25, -0.25) rectangle (7, 0.25);

        \draw [->, postaction={decorate,decoration={raise=2ex, text along path,text align=center,text={stream 3}}}] (2, -1.5) -- (7.5, -1.5);
        \filldraw [fill=purple!20, draw=black] (2.5, -1.75) rectangle (3.25, -1.25);
        \filldraw [fill=purple!20, draw=black] (3.75, -1.75) rectangle (4.5, -1.25);
        \filldraw [fill=purple!20, draw=black] (5, -1.75) rectangle (5.75, -1.25);
        \filldraw [fill=purple!20, draw=black] (6.25, -1.75) rectangle (7, -1.25);


    \end{tikzpicture}
    \caption{Stream multiplexing in a single QUIC connection}
    \label{fig:stream-multiplexing}
\end{figure}

\subsection{Connection migration}
\label{subsec:connection-migration}
Each QUIC connection has a set of connection identifiers each of which can identify the connection.
Connection IDs can be variable length.
QUIC uses connection IDs to route packets to proper endpoint.
This way, as long as there is an alternative network path, QUIC is resilient to network changes avoiding restarting the connection when an endpoint changes its transport address (IP or port).

\subsection{Flow control and congestion control mechanisms}
\label{subsec:flow-control-and-congestion-control}
Flow control limits amount of data that can be sent to the receiver.
Receiver can announce how much data it is able to process/buffer and sender must not exceed this limit until it is updated.
In QUIC, flow control limits can be set on two levels.
The first one is stream level.
This limit is set per stream type e.g.\ for all unidirectional streams opened by the peer or for all bidirectional streams opened locally.
There is no possibility to set flow control limit for a stream with a particular id.
The second one is connection level and it limits overall data amount that can be sent within the whole connection across all streams.
QUIC allows also for limiting number of streams that can be opened by the peer.
These limits are also set per stream type -- unidirectional or bidirectional and initiated by a peer or locally.

On the other hand congestion control limits amount of data that can be sent through the network.
Whenever there is a danger that some middle-boxes like routers can be overwhelmed by the amount of data they have to process, they require sender to slow down its transmission rate.
This is often done using ECN (Explicit Congestion Notification) flag in an IP packet.
RFC 9001~\cite{rfc9001} defines congestion controller similar to TCP NewReno.
However, any other congestion controller that meets requirements defined in section 3.1 of RFC 8085~\cite{rfc8085} can be used.
This makes QUIC congestion control mechanism pluggable.

\subsection{Reliability}
\label{subsec:reliability}
QUIC is a reliable protocol which means it guarantees delivery of each packet sent by the endpoint.
Packets are delivered to the receiver side in the order they were sent.
However, there are different IETF drafts that expand QUIC with new features.
One of them called \textit{An Unreliable Datagram Extension to QUIC}\cite{bider-ssh-quic-09} allows for sending unreliable messages over QUIC\@.
This can take place in simultaneously to the reliable communication.
Unreliable messages are not subject to flow control mechanisms however, they are congestion controlled.
Each unreliable message is also acknowledged so that application layer can be provided with the packet loss information.
Section~\ref{sec:datagrams} describes Datagram extension in details.

\subsection{QUIC against other transport protocols}
\label{subsec:quic_against_other_transport_protocols}
This section compares QUIC with three main transport protocols -- TCP, UDP and SCTP\@.

\subsubsection{TCP}
Transmission Control Protocol -- connection based, stream oriented and reliable transport protocol.
It guarantees the order of messages and provides flow control and congestion control mechanisms.
TCP is not a multiplexed protocol.
In one TCP connection we have one logical communication channel.
This is the reason of head of line blocking problem that appears when many HTTP/2 requests are multiplexed in one TCP connection.
Because of its reliability, TCP is not a good choice for many types of interactive communication.

\subsubsection{UDP}
User Datagram Protocol -- connection less, unreliable transport protocol.
It does not preserve the order of messages.
Datagrams, in this protocol, are not acknowledged nor congestion controlled.
It does not also introduce any flow control mechanisms.
Thanks to its simplicity and low bandwidth affection it is the base for RTP protocol which is widely used for transmitting multimedia.
Unlike TCP, UDP also allows for multicast transmission.
However, multicast is rarely used, mostly in local networks.

\subsubsection{SCTP}
Stream Control Transmission Protocol -- reliable, based on UDP, message oriented transport protocol.
Unlike TCP, SCTP is multistream protocol.
SCTP is partially ordered but it also allows for sending unordered messages.
Another feature of SCTP is multi-homing in which an endpoint uses two different addresses.
This introduces some kind of fault tolerance -- one address is designated as a primary and the other one can be used in case of failure of the first one.
Unfortunately SCTP is not widely used.
Its main domain is telecommunication as a lot of home routers does not handle it properly.

\subsubsection{Summary}
Table~\ref{tab:protocols_comparision} presents a short summary of transport protocols comparison.
\begin{table}
    \centering
    \resizebox{\columnwidth}{!}{%
    \begin{tabular}{|c | c | c | c | c |}
        \hline
        & TCP           & UDP              & SCTP              & QUIC          \\
        \hline
        reliability        & reliable      & unreliable       & reliable          & reliable      \\
        \hline
        transmission       & byte oriented & message oriented & message oriented  & byte oriented \\
        \hline
        flow control       & yes           & no               & yes               & yes           \\
        \hline
        congestion control & yes           & no               & yes               & yes           \\
        \hline
        packets order      & ordered       & unordered        & partially ordered & ordered       \\
        \hline
        multistream        & no            & no               & yes               & yes           \\
        \hline
    \end{tabular}%
    }
    \caption{\label{tab:protocols_comparision}Transport protocols comparison.}
\end{table}
